Towards an Ecological Trait-data Standard

### 2018-09-10 manuscript/manuscript.Rmd
revise MS for comments of TRY and OpenTraits

1. Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions. Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a general and simple vocabulary as well as a simple data structure for storing and sharing ecological trait data. 
4. Additionally, we provide an R-package that enables the transformation of any tabular dataset into the proposed format. This also allows trait datasets from heterogeneous sources to be harmonised and merged, thus facilitating data compilation for any particular research focus. 
5. With these decentralised tools for trait-data harmonisation, we intend to facilitate the exchange and analysis of trait data within ecological research and enable global syntheses of traits across a wide range of taxa and ecosystems.

@@

Many issues in trait-based research arise when compiling datasets from several sources. Data may differ in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements. These differences are not always documented in the metadata accompanying a dataset. 
All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, the number of available trait datasets is increasing rapidly. 
In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
<!--These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated --> 
Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a minimal structure and vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. Finally, we present an R package that assists the harmonisation of trait data from distributed sources. 
With this easy-to-use terminology and toolset, we hope to  convince trait-data providers and trait-data users about the general importance of trait-data standardisation and lay out the roadmap towards an accessible ecological trait data standard.

@@

FDS: see Garnier et al. but may contain:

-->

@@

In the field of comparative biology, morphological traits related to plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values for plants or ecological parameters of animals (e.g. reproductive traits, feeding biology, dispersal or body size) have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
Today, the online publication of such data is greatly facilitated by  *file hosting services* (e.g. Figshare.com), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low thresholds for *metadata* documentation and data standardisation. 
<!-- As a result, datasets greatly vary in structure and terminology. -->

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for a current excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.  
As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Appendix A table A1].<!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static data tables on low-threshold file hosting platforms and are updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with standardisation. 
Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available.

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created to cover certain trait types (e.g. floral traits, seed traits, root traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g. location or climatic data).
Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret OTUs in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, initiatives for particular groups of animals, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/), already exist.

Regarding open access, few of these centralised databases comply with the criteria demanded by journals and funding agencies for primary data publication. The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship.
With the proactive turn towards open access data (as stated in the Bouchot Declaration; http://www.bouchoutdeclaration.org/), it may be necessary to find other incentives for data submission.

@@

These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and even cover most use cases of trait data. An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

@@

By filling this gap, data-brokering services [e.g. the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/; ) are likely to gain importance. These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->

@@

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation. 
Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait datasets, but data harmonisation of these distributed data sets is currently laborious. 
The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
We propose the development of  tools and vocabularies that impose low thresholds and offer high pay-off in the visibility and interconnectedness of published data.

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research.
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the semantic web.
By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to the work of @kattge11, @kattge11a, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases. 
Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. 
The vocabulary proposed is intended to form the foundations of a standard nomenclature that can be expanded and corrected by the wider community of researchers using trait-based approaches in ecology.
<!--
## Why use the vocabulary

@@

## How to apply the standard vocabulary

We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible.
The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observed on an individual or population (i.e. entity), of a biological taxon.
When applying the vocabulary, it is implicit to use a two-dimensional observation long-table format for the data (Fig. 2 *b*), rather than a species$\times$traits matrix (Fig. 2 *a*). As the long-table format draws from a defined set of columns, merging datasets is easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14].

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g. "Schneider, F.D., Jochum, M.,  Le Provost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2018 Ecological Traitdata Standard v0.8, DOI:  10.5281/zenodo.1255287, URL: https://ecologicaltraitdata.github.io/ETS/v0.8/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e. via the API; **in preparation!**). 
Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
The core terms describe minimal trait data according to the Entity-Quality model. 
Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information). These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below) which may simply be added as extra columns to the core dataset.
Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below). 
The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

@@

Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets (published separately or as part of a Darwin Core Archive), thus avoiding redundancy and duplication of content. A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken [@groom17; @guntsch17].

@@

# Computational tools for producing compliant data

>>>>

1. Trait-based approaches are widespread throughout ecological research, offering great potential for trait data to deliver general and mechanistic conclusions. Accordingly, a wealth of trait data is available for many organism groups, but, due to a lack of standardisation, these data come in heterogeneous formats.
2. We review current initiatives and infrastructures for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories.
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets, we propose a standardised vocabulary to be used for storing and sharing ecological trait data. 
4. This vocabulary allows trait datasets from heterogeneous sources to be merged more easily into databases, and facilitates the creation of data interfaces between software tools for trait-data handling and analysis. 
5. By enabling decentralised trait-data standardisation, we intend to facilitate the exchange and analysis of trait data for a broader ecological research community and enable global syntheses of traits across a wide range of taxa and ecosystems.

@@

Many issues in trait-based research arise when compiling datasets from multiple sources. Data may differ in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements. These differences are not always documented in the metadata accompanying a dataset. 
All of these factors render trait data extremely heterogeneous and make the task of data compilation time-consuming or even prohibitive.
However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa ase well as in evolutionary deep-time.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, the number of available trait datasets is increasing rapidly. 
In the past, trait data have been standardised and compiled in centralised databases for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. 
<!--These databases are then published as static data tables (i.e. product of a one-time effort as it is the case of AmphiBIO), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods (e.g. TRY, www.try-db.org). Furthermore,--> These initiatives map heterogeneous data into a common scheme and, importantly, may also offer access control and data usage policies. As such, they protect the rights of the original data providers while simplifying data queries for synthesis researchers. 
Besides initiatives aiming at assembling data, other tools to enable the compatibility of data across databases are being developed. These include semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17].
<!-- Solving the numerous issues of data assemblage and harmonisation in centralised infrastructures requires a huge effort in costs and personnel. -->
<!-- 4. distributed trait data must be integrated -->

Meanwhile, open-science reaches the mainstream: it has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. As a result, an increasing number of individual research projects publish their primary data on file hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org, where no data standards are forced upon the uploaded material. <!-- The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.-->
It is likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient re-use and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, as well as initiatives for standardisation. We discuss current practice and the importance of data standards for trait-based research, and we identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we propose a versatile vocabulary for describing trait datasets, that builds upon and is compatible with existing terminology standards for biodiversity data. 
Since this standard vocabulary relies on the adoption of a broad research community we discuss incentives for its use and lay out mechanisms for a community-based future development towards an accessible ecological trait data standard.
<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->

@@

FDS: see Garnier et al. but may contain: 
-->

@@

In the field of comparative biology, morphological traits related to plant flower, leaf and stem traits or bird wing and beak measurements, as well as life-history traits such as Ellenberg values for plants or ecological parameters of animals (e.g. reproductive traits, feeding biology, dispersal or body size) have been measured for decades, and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, individual measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published along with scientific articles or as standalone data publications [see @kleyer08 for a review on plant data; on animal data, see e.g., @gossner15; @ricklefs17].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->
Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g. Figshare.com, datadryad.org, Researchgate.net, or Zenodo.org), which warrant long-term accessibility and citability via DOIs, and support Public Domain dedication or Creative Commons licenses.
These platforms offer publicly accessible repositories at low-cost or for free, which makes them attractive for small and intermediate sized research projects that cannot dedicate extra resources for data management. 
However, although open for manual access, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low thresholds regarding *metadata* documentation and data standardisation. As a result, published datasets greatly vary in structure and nomenclature.

@@

A global overview of existing trait data for all taxa and trait types is difficult to obtain. Therefore, in an attempt to collate a list of existing distributed datasets, we initiated a living spreadsheet (https://goo.gl/QxzfHy) which lists published trait datasets, their regional and taxonomic focus, the number and scope of traits covered, their location on the internet and the terms of use (see Appendix A for an excerpt of this list). We invite data owners and users to add further trait datasets to this spreadsheet.  
As it stands, the decentralisation and the lack of data standardisation of low-threshold online repositories renders the compilation of data into larger collections inefficient and reduces the potential of many published datasets to be re-used and combined into broad synthesis analysis.

@@

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @klotz02; @iversen17; @kleyer08; @myhrvold15; @kissling14; @jones09; @oliveira17, see Appendix A table A1].<!--[@wright04; ] -->
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently related to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus of the initiative. 
Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms and may be updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
Concerning organism traits, with the digitisation efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowd-sourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

<!-- database backend, selective download, user access management --> 
More specialised trait-database platforms have been created to cover certain trait types (e.g. floral traits, seed traits, root traits or wood density traits), interaction types (e.g. pollination traits or feeding relationships), or a specific environmental and experimental context of the trait observation (e.g. location or climatic data).
Such database initiatives attract data submissions from a defined research field and take care of the harmonisation process and thereby greatly facilitate data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section).

Microbial ecologists also make frequent use of trait-based approaches to assess genomic function and describe functional diversity at the community level [@krause14; @fierer14;  @fierer12]. Here, 'operational taxonomic units' (OTUs) are derived from metagenomic analysis [@torsvik02; @langille13]. Databases are also used to interpret OTUs in terms of their functional role [e.g. the KEGG orthology, @kanehisa12]. 
For animals, a single unified platform and harmonising scheme for animal trait data is still lacking. The reason for this may be that harmonising trait data on animals, which span multiple trophic levels and possess diverse body plans, is a more complex task than for plants [@moretti17]. Nonetheless, initiatives for particular groups of animals do exist, such as the BETSI database collects traits on soil invertebrates [http://betsi.cesab.org/; @pey14], and the Carabids.org web portal collects traits of carabid beetles (http://www.carabids.org/).

The platforms incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and potentially co-authorship. 
However, the upload of data to these centralised databases does currently not fulfill the criteria demanded by journals and funding agencies for primary data publication with long-term availability, citable DOIs and no access restrictions. With data being publicly available, the role of database initiatives in standardising data and making them more accessible is paramount.

@@

These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @holetschek12] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and even cover most use cases of trait data. An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12]. 
For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies the terms described therein.
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

@@

By filling this gap, data-brokering services [e.g. the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/; Diversity Workbench, www.diversityworkbench.net; BExIS, http://bexis2.uni-jena.de/; ) are likely to gain importance. These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data usage in publications and provide standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. Another solution for data-users to access trait data in a structured way is offered by FENNEC [@ankenbrand18], an online tool or self-hosted service to extract trait information from multiple sources for a target species community.  <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->

@@

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation. 
On the other end, public repositories on file-hosting services offer the publication of scientific primary data according to current funding and journal criteria, attracting a wealth of heterogeneous trait datasets. Due to the low thresholds concerning data standardisation, compilation of these distributed data sets is currently laborious. 
The goal must be to better integrate these distributed data into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
We propose the development of tools and vocabularies that impose low thresholds and offer high pay-off in the visibility and interconnectedness of published data.

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research.
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases:

1. in  two-dimensional datasets, e.g. for labelling project-specific data, for local use and exchange with collaborators, or for publication of standardised primary data on open-access data repositories, 
2. as a consensus vocabulary when compiling data from distributed sources into databases, e.g. to map standardised columns and refer to taxa and trait definitions in a uniform way, and
3. as an exchange format of data between web services and software tools, e.g. when providing database queries from a webservice or defining input and output formats of software packages.

By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. 
Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases. 
<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
The vocabulary proposed is intended to form the foundations of a standard exchange format that can be expanded and revised by the wider community of researchers using trait-based approaches in ecology.
<!--
## Why use the vocabulary

@@

@@

The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observed on an individual or population (i.e. entity), of a biological taxon.
Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information). 
These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see below for best practice when publishing primary data).
Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).

Some of the terms expand and refine terms of the Darwin Core Extensions (http://rs.tdwg.org/dwc/terms/) or Glossary of EOL (http://eol.org/info/516). Wherever our glossary refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns. However, these should be clearly specified in the meta-information accompanying the dataset. 
The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

@@

Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content. A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken [@groom17; @guntsch17].

@@

## Best practice for primary data publication

We focus this section on the first of the three use-cases for the ecological trait-data standard vocabulary: The publication of project-specific primary data.  We assume that data managers and software developers who apply the vocabulary in use-case 2 and 3 are more fluent with the use of exchange formats and standard definitions and have established own best-practice principles.
